---
title: "R Package Survey"
author: "Seth Russell"
date: "6/19/2018"
output: html_document
---

This notebook performs an analysis of use of testing methodologies for all R packages avaialble on CRAN.

Two methods are used and results are compared.

*Method 1:* Grep for non-empty testing directories. The commonly used R testing packages all recommend placing tests in a directory by themselves, which we look for.

*Method 2:* Check for stated dependencies on testing packages. It is considered best practice to list dependencies on a testing framework even though standard usage of a package may not require it.


I've Picked 2008 (10 years) as the last updated cutoff year based on manual inspection. While there are still packages in CRAN that were last updated in 2007 and prior, there are not very many, so ignore them in the visualizations.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringi)
library(magrittr)
library(stringr)
library(rvest)
library(ggplot2)

# Testing packages
test_packages <- c("svUnit", "RUnit", "testthat", "testit", "unitizer", "unittest")
# testit is a minimal tool for testing packages, but it can be used for unit testing - e.g. tinytex
# Packages considered but excluded:
#   webmockr - Could be used for a testing library - or perhaps a roll your own test suite, but isn't as far as I could tell
#   vdiffr - Is a extension for testthat, so requires testthat to work
#
# Users could use core R functionality for unit tests such as stopifnot() instead of packages
# we don't attempt to quantify the scope of software that does unit testing without relying on
# a package
```

## Method 1 of testing analysis

This method downloads packages from CRAN and evaluates the files and folders for the presence of a non-empty test directory.

```{r download_files}
# TO DO
# ?? maybe use map instead of lapply ??
# probably want vector of logicals, but need to remove nested lists
# need to evaluate all the packages that don't use the above test directory
# do they use builtin R functions for testing? Or, some older testing packages?
# convert to Notebook
# generate results for paper
# next - look at optimization
#   percent of packages using C, Rcpp, C++, Fortran/Cobol?


read_from_cran <- TRUE
download_dir <- '../downloads'
if(! file.exists(download_dir)) {
  dir.create(download_dir)
}
setwd(download_dir)

# if value 100%, then read all packages, otherwise, randomly select number of packages provided
sample_size <- '100%'
sample_size <- '5'
# dont need to untar, but is useful for manual analysis
untar_files <- TRUE

if (read_from_cran) {
  # canonical source CRAN URL
  # url_base <- "https://cran.r-project.org/src/contrib/"
  # CRAN mirror
  url_base <- "https://cloud.r-project.org/src/contrib/"
  html_source <- read_html("https://cran.r-project.org/src/contrib/")

  td <- html_source %>% html_nodes("td") %>% html_text(trim = TRUE)

  n <- length(td)
  package_df <- data.frame(
    #name
    td[seq(2, n, 5)],
    #package
    stri_replace_last(td[seq(2, n, 5)], '' , regex = '_[0-9.-]+\\.tar\\.gz'),
    # date
    td[seq(3, n, 5)],
    # size
    td[seq(4, n, 5)]
  , stringsAsFactors=FALSE)
  names(package_df) <- c('filename', 'package', 'date', 'size')

  #only want to keep R packages, not other stuff available
  package_df <- package_df[grep(".tar.gz", package_df$filename),]

  # this line for 100% sample
  if(sample_size == '100%') {
    s_size <- nrow(package_df)
  }
  else {
    s_size <- sample_size
  }
    
  files_and_dirs <- lapply(sample(package_df$filename, s_size), function(s) {
    if(file.exists(s)) {
      print(paste0('File ', s, 'already downloaded'))
    } else {
      print(paste0('Attempting to download: ', s))
      tryCatch(download.file(paste0(url_base, s), s),
               error=function(e) NULL)
    }
    if(untar_files) {
      untar(s)
    }
    untar(s, list=TRUE)
  })
} else {
  packages <- list.files(getwd(), pattern=".tar.gz")
  files_and_dirs <- lapply(sample(packages, sample_size), function(s) {
    print(s)
    if (untar_files) {
      untar(s)
    }
    #paths_with_test <- lapply(untar(s, list=TRUE), grepl, pattern="[Tt]est[^/]*/.+", value=TRUE)
    v <- any(grepl(pattern="[Tt]est[^/]*/.+", untar(s, list=TRUE)))
    names(v) <- s
    v
  })
}
files_and_dirs <- unlist(files_and_dirs)

# example of grep stmts to make sure it is working as expected
any(grepl("[Tt]est[^/]*/.+", c('blah', 'blah2')))
any(grepl("[Tt]est[^/]*/.+", c('unitest/myfile.R', 'something')))
any(grepl("[Tt]est[^/]*/.+", c('unitest/f.R', 'tests/b.R')))

# now, for each dirs, see if there is a directory or file named test
# find paths that have test in them
# Although most unit testing packages use test/ or tests/ as the common base directory, some
# such as svUnit recommend the directory /inst/unitTests
#
# code at https://github.com/rorynolan/exampletestr/blob/master/analysis/CRAN_test_analysis.Rmd misses some
# test library usage since it checks for "^tests/.+". As an example, R packages using svUnit (such as gsubfn)
# often call the test directory unitTests
#
paths_with_test <- unlist(lapply(files_and_dirs, grep, pattern="[Tt]est[^/]*/.+", value=TRUE))
# get name of package from base of path
packages_with_tests <- unique(unlist(lapply(paths_with_test, stri_extract_all_regex, pattern="^\\w+[^/]+")))

package_df$uses_tf <- FALSE
package_df[package_df$package %in% packages_with_tests, ]$uses_tf <- TRUE

```

### Visualizations for method 1

Histogram by year with testing packages shown

```{r grep_based_graphics}
# histogram packages that are tested by most current year package released
grep_freqs <- aggregate(strftime(package_df$date, "%Y"),
                   by=list(strftime(package_df$date, "%Y"), package_df$uses_tf),
                   FUN=length)
names(grep_freqs) <- c('Year', 'Uses_TF', 'Count')
grep_plot <- ggplot(data=grep_freqs[grep_freqs$Year > 2007, ], aes(x=Year, y=Count, fill=Uses_TF)) +
  geom_bar(stat="identity")
grep_plot
ggsave(filename = "../sourcecode/file_analysis_stacked_bar.png", grep_plot,
       width = 7.2, height = 5.5, dpi = 600, units = "in", device='png')

# plot (table?) showing tested as pct of all packages
grep_totals <- aggregate(x=grep_freqs$Count, by = list(grep_freqs$Year), FUN = sum)
names(grep_totals) <- c('Year', 'Count')
grep_totals <- merge(grep_totals, grep_freqs[grep_freqs$Uses_TF == TRUE, ][c('Year', 'Count')], by="Year")
names(grep_totals) <- c('Year', 'Total', 'Grep')
grep_totals$Pct <- grep_totals$Grep / grep_totals$Total

grep_totals

grep_pct_plot <- ggplot(data=grep_totals[grep_totals$Year > 2007, ], aes(x=Year, y=Pct * 100)) +
  geom_bar(stat="identity") +
  ylab("Percent with Testing Code") +
  xlab("Year Package Last Updated") +
  scale_y_continuous(limits = c(0,100))
grep_pct_plot
ggsave(filename = "../sourcecode/pct_w_testing_code.png", grep_pct_plot,
       width = 7.2, height = 4, dpi = 600, units = "in", device='png')

```

## Method 2 of testing analysis

```{r dependencies}
# Calculate which testing packages are most popular
# reverse dependencies don't give the whole story - e.g. Wats depends on testit, but as shown
# in the directory structure, uses testthat for unit testing
#
# since package source not provided, using cran.r-project.org by default
local({r <- getOption("repos")
       r["CRAN"] <- "https://cran.rstudio.com/" 
       options(repos=r)
})
test_package_dependencies <- tools:::package_dependencies(packages = test_packages,
                                                          recursive=FALSE,
                                                          reverse=TRUE,
                                                          which = c("Depends","Imports","LinkingTo", "Suggests"))
# Need check for use of various testing frameworks
#   - count how many use each
#   - count total number of packages using a testing package
package_df$framework <- 'none'
print(sort(unlist(Map(length, test_package_dependencies)), decreasing = TRUE))
result <- Map(function(tpd) {
  package_df[tpd] <<- FALSE
  Map(function(p) {
    package_df[package_df$package == p, ][tpd] <<- TRUE
    package_df[package_df$package == p, ]$framework <<- tpd
  }, test_package_dependencies[[tpd]])
}, names(test_package_dependencies))

# are there any that have more than one testing framework listed?
# these are all counted multiple times in histogram
package_df[rowSums(package_df[test_packages]) > 1, ]
nrow(package_df[rowSums(package_df[test_packages]) > 1, ])

```

### Visualizations for method 2

Histogram by year with testing packages shown

```{r dependency_based_graphics}
dep_freqs <- aggregate(strftime(package_df$date, "%Y"),
                   by=list(strftime(package_df$date, "%Y"), package_df$framework),
                   FUN=length)
names(dep_freqs) <- c('Year', 'Framework', 'Count')
fplot <- ggplot(data=dep_freqs[dep_freqs$Year > 2007, ], aes(x=Year, y=Count, fill=Framework)) +
  geom_bar(stat="identity")
fplot
ggsave(filename = "../sourcecode/framework_stacked_bar.png", fplot,
       width = 7.2, height = 5.5, dpi = 600, units = "in", device='png')

dep_totals <- aggregate(x=dep_freqs$Count, by = list(dep_freqs$Year), FUN = sum)
names(dep_totals)[1] <- 'Year'
Map(function(t) {
  dep_totals <<- merge(dep_totals, dep_freqs[dep_freqs$Framework == t, ][c('Year', 'Count')], by="Year", all=TRUE)
  names(dep_totals)[length(names(dep_totals))] <<- t
  print(dep_totals)
  dep_totals[paste0(t, '_Pct')] <<- dep_totals$t / dep_totals$x
}, test_packages)
dep_totals[is.na(dep_totals)] <- 0
names(dep_totals)[2] <- 'Count'
dep_totals

# testthat is the only one that has percentages above 1, so only plot that
dep_pct_plot <- ggplot(data=dep_totals[dep_totals$Year > 2007, ], aes(x=Year, y=testthat_Pct * 100)) +
  geom_bar(stat="identity") +
  ylab("Percent depending on testthat") +
  xlab("Year Package Last Updated") +
  scale_y_continuous(limits = c(0,100))
dep_pct_plot
ggsave(filename = "../sourcecode/pct_w_testthat_dep.png", dep_pct_plot,
       width = 7.2, height = 4, dpi = 600, units = "in", device='png')


```

### Method comparison

As the results from method 1 do not match the results from method 2, explore some of the differences.

Some points discovered:
* It is possible for an R package to use a non-standard testing directory, even if they also use a standardized testing framework
* It is possible to list a testing framework as a dependency and not acutally use the framework.
* Speculation - Testing can be performed in-line with other code through use of stop() or stopifnot().
* Speculation - Developers may have their own test cases they run while developing their software, but do not commit to a repository or share with others.

```{r comparison}
nrow(package_df[package_df$uses_tf == TRUE, ]) # 851 more appear to have testing code...
nrow(package_df[package_df$framework != 'none', ])

# 1062 packages that don't list a dependency to a testing framework, but have a testing directory
# Randomly looked at several: some don't use a framework (e.g. xlsx)
# others use a framework but don't list it as a dependency (e.g. redcapAPI)
only_grep <- package_df[package_df$uses_tf == TRUE & package_df$framework == 'none', ]
nrow(only_grep)

# 211 packages that list a dependency to a testing framework, but don't use one
only_dep <- package_df[package_df$uses_tf == FALSE & package_df$framework != 'none', ]
nrow(only_dep)

```


